---
title: "Project3_bmi_trajectory_MinjiKim"
output: html_document
date: "2024-11-19"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## setting
```{r}
#install.packages("rjags", type = "source")
#install.packages("JM", type = "source") # for joint model
library(tidyverse)
library(haven)
library(janitor)
library(survival)
library(survminer)
library(gtsummary)
library(broom)
library(lubridate)

library(ggplot2)
library(tidyverse)
library(lme4)
library(lmtest)
library(patchwork)
library(splines)
library(nlme)
library(clubSandwich)
library(rjags)
library(JM)
library(lmerTest)
library(dplyr)
library(lubridate)

# long dataset for logitudinal analysis
df_bmi <- readRDS("post_bmi_studypopulation_2ndanalysis.rds")
# time-to
df <- readRDS("studypopulation.rds") # 112858
```
# cleaning
```{r}
# Ensure bmi_measurement_date is in Date format
df_bmi <- df_bmi %>% mutate(bmi_date = as.Date(bmi_date))

# Sort data by patient_id and date
df_bmi <- df_bmi %>%
  arrange(patient_id, bmi_date) %>%
  group_by(patient_id) %>%
  mutate(
    time_diff_months = as.numeric(difftime(bmi_date, lag(bmi_date), units = "days")) / 30.44,
    bmi_diff = bmi - lag(bmi),
    bmi_change_per_month = bmi_diff / time_diff_months
  ) %>%
  ungroup()

# Exclude outliers where BMI change exceeds 4 per month
df_bmi <- df_bmi %>%
  filter(is.na(bmi_change_per_month) | abs(bmi_change_per_month) <= 5)

# Print summary of removed outliers
bmi_outliers <- df_bmi %>%
  filter(!is.na(bmi_change_per_month) & abs(bmi_change_per_month) > 5)

cat("Total Outliers Removed:", nrow(bmi_outliers), "\n") # 0

# remove the covariates
df_bmi <- df_bmi %>% dplyr::select(- time_diff_months, - bmi_diff, - bmi_change_per_month, -bmi_index, -glp1_initiation_date, -death_date)

df_bmi <- df_bmi %>% filter(!is.na(bs_type)) # remove 15 individuals 

df_bmi <- df_bmi %>% rename(glp1 = glp1_user)
# exploring outcome variable (BMI)
hist(df_bmi$bmi)

# check missingness
mice::md.pattern(df_bmi, plot=FALSE)
```


# 1. Longitudinal data analysis component

Model: linear mixed model with natual spline term
Data frame: long format

### 1.1. data preprocessing 
```{r}
# make continous time variable and indicator for exposure
df_bmi <- df_bmi %>% mutate(time = bmi_date - bs_date)
df_bmi$time <- as.numeric(df_bmi$time)
df_bmi$glp1 <- as.factor(df_bmi$glp1)
df_bmi$time_to_glp1_cat <- as.factor(df_bmi$time_to_glp1_cat)

# remove unappropriate data - BMI < 10
df_bmi <- df_bmi %>% filter(bmi >= 10)

# remain BMI only after index date
df_bmi <- df_bmi %>% filter(bmi_date >= bs_date)
nrow(df_bmi) # 1115993
```
### ### for removing the sudden drops, 
For every current BMI, calculate {(current BMI – past BMI)/number of days between the two observations} * 365.25 and remove any current BMI that has greater absolute yearly change than x%, which depends on how you define an outlier. You can start with 75% - 100% and see if outliers disappear from the spaghetti plot and fit the models again.

### remove individuals have only one BMI data point.
```{r}
# assess how many individuals have only one BMI data point?
individual_counts <- df_bmi %>%
  group_by(patient_id) %>%
  summarise(row_count = n()) %>%  # Count rows per individual
  filter(row_count == 1)
num_individuals <- nrow(individual_counts)
cat("Number of individuals with only one row:", num_individuals, "\n") # 736 

# remove that individuals
df_bmi <- df_bmi %>%
  group_by(patient_id) %>%
  filter(n() > 1) %>%  # Keep only individuals with more than one row
  ungroup()            # Remove grouping for further operations
nrow(df_bmi) # 1115257
```
### set several dataset for BMI measurements
```{r}
# with at least 3 points of BMI
# remove that individuals
df_bmi <- df_bmi %>%
  group_by(patient_id) %>%
  filter(n() > 1) %>%  # Keep only individuals with more than one row
  ungroup() 
```

## OUTLIERS
### calculate the mean and SD by individual - don't proceed
```{r}
df_bmi <- df_bmi %>%
  group_by(patient_id) %>%
  mutate(
    mean = mean(bmi),
    sd = sd(bmi)
  )
head(df_bmi)

outliers <- df_bmi %>%
  group_by(patient_id) %>%
  filter(
    bmi > mean(bmi, na.rm = TRUE) + 5 * sd(bmi, na.rm = TRUE) |
    bmi < mean(bmi, na.rm = TRUE) - 5 * sd(bmi, na.rm = TRUE)
  )
nrow(outliers) # 97 with 5sd | 391 with 4 sd | 1890 with 3sd
unique_patients <- length(unique(outliers$patient_id))
print(unique_patients) # 89 individuals with 5sd | 337 with 4 sd | 1426 with 3sd

89/15606*100  # 0.57%
337/15606*100  # 2.16%
1426/15606*100  # 9.14%

# Extract unique patient IDs
outliers_patients <- unique(outliers$patient_id)

# Filter original dataset to include only outlier patients
df_bmi_outliers <- df_bmi %>%
  filter(patient_id %in% outliers_patients)

unique_patients <- length(unique(df_bmi_outliers$patient_id))
print(unique_patients) # 89
head(df_bmi_outliers)
```
## outlier definition with more than 5 BMI per month
```{r}
head(df_bmi)

# Ensure bmi_measurement_date is in Date format
df_bmi <- df_bmi %>% mutate(bmi_date = as.Date(bmi_date))

# Sort data by patient_id and date
df_bmi <- df_bmi %>%
  arrange(patient_id, bmi_date) %>%
  group_by(patient_id) %>%
  mutate(
    time_diff_months = as.numeric(difftime(bmi_date, lag(bmi_date), units = "days")) / 30.44,
    bmi_diff = bmi - lag(bmi),
    bmi_change_per_month = bmi_diff / time_diff_months
  ) %>%
  ungroup()

# Exclude outliers where BMI change exceeds 4 per month
df_bmi_clean <- df_bmi %>%
  filter(is.na(bmi_change_per_month) | abs(bmi_change_per_month) <= 5)

# Print summary of removed outliers
df_bmi_outliers <- df_bmi %>%
  filter(!is.na(bmi_change_per_month) & abs(bmi_change_per_month) > 5)

cat("Total Outliers Removed:", nrow(df_bmi_outliers), "\n")

# View the cleaned dataset
head(df_bmi_clean)
```

## make spagetti plot with the outliers
```{r}
df_bmi_outliers

# plot with random subset
set.seed(122)
id_sel = sample(df_bmi_outliers$patient_id, 1)

# total
ggplot() +
  geom_point(data=df_bmi_outliers[df_bmi_outliers$patient_id %in% id_sel, ], aes(x=time, y=bmi), alpha=0.7) +
  geom_line(data=df_bmi_outliers[df_bmi_outliers$patient_id %in% id_sel, ], aes(x=time, y=bmi, group=patient_id), alpha=0.5) +
  theme_classic() +
  ggtitle("Post-surgical BMI of patient with outliers [awCFE] ")  + 
  xlab("Time (in year)") +
  ylab("BMI") + 
  scale_x_continuous(
    breaks = seq(0, max(df_bmi$time_n, na.rm = TRUE), by = 365), 
    labels = seq(0, max(df_bmi$time_n, na.rm = TRUE) / 365, by = 1)
    )

awCFE <- df_bmi %>% filter(patient_id == "awCFE")
awCFE_out <- awCFE %>% filter(bmi < 20)
view(awCFE)
```
### remove the row with the outliers from the original dataset
```{r}
df_bmi <- df_bmi %>%
  anti_join(df_bmi_outliers, by = c("patient_id", "bmi", "bmi_date"))
nrow(df_bmi)

# remove the covariates
df_bmi <- df_bmi %>% dplyr::select(- time_diff_months, - bmi_diff, - bmi_change_per_month)
```
### check
```{r}
# number of study population # 15618 <- glp1 users = 2460 | non-users = 13158
unique_patients <- length(unique(df_bmi$patient_id[df_bmi$glp1 ==0]))
print(unique_patients)

# seperate database based on the GLP-1 user vs. non-users
df_bmi_glp1 <- df_bmi %>%
  filter(glp1 ==1) 

df_bmi_non <- df_bmi %>%
  filter(glp1 ==0)
```

### categorized
```{r}
table(df$time_to_glp1_cat)

# use all 8 categories
a <- df_bmi %>%
  distinct(patient_id, time_to_glp1_cat)
nrow(a)
table(a$time_to_glp1_cat)

df_bmi$time_to_glp1_cat <- as.factor(df_bmi$time_to_glp1_cat)

# indicators : never | early | middle | late
# time_to_glp1_cat2 == 0 for Never use (n=13158)
# time_to_glp1_cat2 == 1 for Early use (within 2 years, n=511)
# time_to_glp1_cat2 == 2 for Middle use (between 2-5 years, n=1182)
# time_to_glp1_cat2 == 3 for Late use (after 5th years, n=767)

df_bmi <- df_bmi %>% 
  mutate(
    time_to_glp1_cat2 = case_when(
      time_to_glp1_cat == 0 ~ 0,
      time_to_glp1_cat %in% c(1, 2) ~ 1,
      time_to_glp1_cat %in% c(3,4,5) ~ 2,
      time_to_glp1_cat %in% c(6,7,8) ~ 3
    )
  )
print(df_bmi %>% dplyr::select(patient_id, time_to_glp1_cat,  time_to_glp1_cat2), n=10)

df <- df %>% 
  mutate(
    time_to_glp1_cat2 = case_when(
      time_to_glp1_cat == 0 ~ 0,
      time_to_glp1_cat %in% c(1, 2) ~ 1,
      time_to_glp1_cat %in% c(3,4,5) ~ 2,
      time_to_glp1_cat %in% c(6,7,8) ~ 3
    )
  )
print(df %>% dplyr::select(patient_id, time_to_glp1_cat,  time_to_glp1_cat2), n=10)

df_bmi$glp1 <- as.factor(df_bmi$glp1)
df_bmi$time_to_glp1_cat <- as.factor(df_bmi$time_to_glp1_cat)
df_bmi$time_to_glp1_cat2 <- as.factor(df_bmi$time_to_glp1_cat2)

# bar graph for the time_to_glp1_cat
ggplot(a %>% dplyr::filter(time_to_glp1_cat != 0), aes(x = time_to_glp1_cat)) +
  geom_bar(fill = "blue") +
  geom_text(aes(label = ..count..), stat = "count", vjust = -0.5, colour = "black") +
  theme_classic() +
  labs(
    x = "Time to GLP-1 Category",
    y = "Count",
    title = "Distribution of Time to GLP-1 Categories among GLP-1 users"
  ) + 
  scale_x_discrete(
    breaks = c(1,2,3,4,5,6,7,8),  # Specify positions of the intervals
    labels = c("within 1 year", "in 1-2 year", "in 2-3 year", "in 3-4 year", "in 4-5 year", "in 5-6 year", "in 6-7 year", "after 7 year")  # Custom labels
  ) 
```

## add spline term for the last one year
```{r}
# set the event date | censored date
# 1. glp1 users = glp1_initiation_date
df_bmi_glp1 <- df_bmi %>% filter(glp1 ==1)

df_bmi_glp1 <- df_bmi_glp1 %>%
  left_join(
    df %>% dplyr::select(patient_id, glp1_initiation_date), 
    by ="patient_id"
  ) %>%
  rename(date_censored = glp1_initiation_date)

# 2. non-users = the last bmi date
df_bmi_non <- df_bmi %>% filter(glp1 ==0)

df_bmi_non <- df_bmi_non %>%
  group_by(patient_id) %>%
  mutate(
    date_censored = max(bmi_date)
  ) %>% ungroup()

# 3. merge two files
df_bmi <- rbind(df_bmi_glp1, df_bmi_non)

# 4. set new time 0
df_bmi <- df_bmi %>%
  mutate(new_time0 = date_censored - 365)

# 5. make a indicators from new time 0
df_bmi <- df_bmi %>%
  group_by(patient_id) %>% 
  dplyr::mutate(
    time_1yr = ifelse(bmi_date >= new_time0, bmi_date - new_time0, 0)
  )
head(df_bmi %>% dplyr::filter(glp1==1) %>% dplyr::select(patient_id, bmi_date, date_censored, time_1yr))
```
### 1.2. EDA

From the histogram, we chose not to transform the data to opt for better interpretability, considering the data is not highly skewed to the left.

However, we need to drop the BMI value == 0.

```{r}
# exploring outcome variable (BMI)
hist(df_bmi$bmi)

# check missingness
mice::md.pattern(df_bmi, plot=FALSE)
```
## Preliminary data display

### Spaghetti plot with random subset

In exploratory analysis, spaghetti plot was utilized to assess the within-subject and between-subject variability in BMI.

We plot a Spaghetti plot to visualize our outcome over time.
Based on the plot, we see that there are large heterogeneity among patients, which warrants the use of random intercept and random slope in our model.

Using random intercept and random slope can account for the non-constant variance and correlation between subjects.
 
```{r}
# Spaghetti plot to show the variance within and between participants
set.seed(123)
id_sel = sample(df_bmi$patient_id, 50)
id_sel_glp1 = sample(df_bmi$patient_id[df_bmi$glp1 ==1], 50)
id_sel_non = sample(df_bmi$patient_id[df_bmi$glp1 ==0], 50)

df_bmi$time_jitter=df_bmi$time + 0.5*(runif(length(df_bmi$time)) - 0.5) + 0.2*ifelse(df_bmi$glp1==0,0,1)

spagh_plot1=ggplot() +
  geom_point(data=df_bmi[df_bmi$patient_id %in% id_sel, ],aes(x=time, y=bmi, group=glp1, color=glp1),alpha=0.5) +
  geom_line(data=df_bmi[df_bmi$patient_id %in% id_sel, ],aes(x=time, y=bmi, group=patient_id,color=glp1),alpha=0.4) +
  ggtitle("Spaghetti plot for BMI")  + 
  xlab("Time") +
  ylab("BMI") + 
  scale_color_manual(
  values = c("0" = "black", "1" = "blue"),
  labels = c("Not on GLP-1", "On GLP-1"),
  name = "GLP-1 Status"
)

spagh_plot1

# for glp1 users
spagh_plot_glp1=ggplot() +
  geom_point(data=df_bmi[df_bmi$patient_id %in% id_sel_glp1, ],aes(x=time, y=bmi, group=glp1), color="blue",alpha=0.5) +
  geom_line(data=df_bmi[df_bmi$patient_id %in% id_sel_glp1, ],aes(x=time, y=bmi, group=patient_id,color=glp1), color="blue",alpha=0.4) +
  ggtitle("Spaghetti plot for BMI in GLP-1 users")  + 
  xlab("Time [days]") +
  ylab("BMI") +
  theme_classic()

spagh_plot_glp1

# for non-users
spagh_plot_non=ggplot() +
  geom_point(data=df_bmi[df_bmi$patient_id %in% id_sel_non, ],aes(x=time, y=bmi, group=glp1), color="black",alpha=0.5) +
  geom_line(data=df_bmi[df_bmi$patient_id %in% id_sel_non, ],aes(x=time, y=bmi, group=patient_id,color=glp1), color="black", alpha=0.4) +
  ggtitle("Spaghetti plot for BMI in non-users")  + 
  xlab("Time [days]") +
  ylab("BMI") + 
  theme_classic()

spagh_plot_non
```
### visit plot with random subset
```{r}
# quantify the number of visit
n_visit_glp1 <- df_bmi_glp1 %>%
  group_by(patient_id) %>%
  summarise(
    number_visits = n()
  )
summary(n_visit_glp1$number_visits)

n_visit_non <- df_bmi_non %>%
  group_by(patient_id) %>%
  summarise(
    number_visits = n()
  )
summary(n_visit_non$number_visits)

boxplot(n_visit_glp1$number_visits, n_visit_non$number_visits)

n_visit_glp1_clean <- n_visit_glp1 %>%
  filter(number_visits < 40)
n_visit_non_clean <- n_visit_non %>%
  filter(number_visits < 40)

boxplot(n_visit_glp1_clean$number_visits, n_visit_non_clean$number_visits)

# plot with random subset
set.seed(123)
id_sel = sample(df_bmi$patient_id, 50)
id_sel_glp1 = sample(df_bmi$patient_id[df_bmi$glp1 ==1], 50)
id_sel_non = sample(df_bmi$patient_id[df_bmi$glp1 ==0], 50)

# total
p_all <- ggplot(df_bmi[df_bmi$patient_id %in% id_sel, ], aes(x=time, y=factor(patient_id), color = glp1)) +
    geom_line(aes(group=factor(patient_id)), color="grey") +
    geom_point(aes(shape=factor(glp1), fill=factor(glp1), group=factor(patient_id)), color="gray30", size=1.1, alpha=0.8) +
    labs(title = "BMI measured date by Individual",
         x = "BMI measured date [days]",
         y = "Individual ID",
         shape = "glp1",
         fill = "glp1") +
    theme_bw() +
    theme(axis.text.y = element_blank())
p_all

# GLP-1 users
p_glp1 <- ggplot(df_bmi[df_bmi$patient_id %in% id_sel_glp1, ], aes(x=time, y=factor(patient_id), color = glp1)) +
    geom_line(aes(group=factor(patient_id)), color="grey") +
    geom_point(aes(shape=factor(glp1), fill=factor(glp1), group=factor(patient_id)), color="blue", size=1.1, alpha=0.8) +
    labs(title = "The date of BMI measurement for each individual among GLP-1 users",
         x = "BMI measured date [days]",
         y = "Individual ID",
         shape = "glp1",
         fill = "glp1") +
    theme_bw() +
    theme(axis.text.y = element_blank())
p_glp1

# non-users
p_non <- ggplot(df_bmi[df_bmi$patient_id %in% id_sel_non, ], aes(x=time, y=factor(patient_id), color = glp1)) +
    geom_line(aes(group=factor(patient_id)), color="grey") +
    geom_point(aes(shape=factor(glp1), fill=factor(glp1), group=factor(patient_id)), color="black", size=1.1, alpha=0.8) +
    labs(title = "The date of BMI measurement for each individual among non-users",
         x = "BMI measured date [days]",
         y = "Individual ID",
         shape = "glp1",
         fill = "glp1") +
    theme_bw() +
    theme(axis.text.y = element_blank())
p_non
```
### fit the model

To account for the autocorrelation and heterogeneity between individuals, we fitted a linear mixed effect model (LMM) allowing negative symptom scores to change as a function of time (natural spline, df=3) and the interaction between time (ns, df=3) and treatment, with additional random intercept and random slope for linear time.  

Among linear mixed models and linear and spline models using OLS, we chose linear mixed models with random intercept and random slope based on the AIC value. 

After selecting final logitudinal model, I added baseline covariates. 
(The same set of covariates will be included in the cox model).

We can also do likelihood ratio test to check if addition of ns(time, df=3) and treatment terms significantly improves fit.

## modeling 2 "linear mixed effect model with random intercept and slope" for each group
```{r}
# Standardize continuous variables
df_bmi$time_scaled <- scale(df_bmi$time)
names(df_bmi)

# model 0 - wrong model
model.lmer0 <- lmer(data=df_bmi, bmi ~ glp1 + ns(time_scaled,3):glp1 + (time_scaled | patient_id), REML=FALSE)
summary(model.lmer0)

# model 1
model.lmer1 <- lmer(data=df_bmi, bmi ~ ns(time_scaled,3) + (time_scaled | patient_id), REML=FALSE)
summary(model.lmer1)

# model 2
model.lmer2 <- lmer(data=df_bmi, bmi ~ ns(time_scaled,3) + bs_type + sex + race + ethnicity + marital_status + patient_regional_location + age_at_bs + cc_t2db + cc_htn + cc_dyslip + cc_hf + cc_liver + (time_scaled | patient_id), REML=FALSE)
summary(model.lmer2)

# model 3
model.lmer3 <- lmer(data=df_bmi, bmi ~ ns(time_scaled,3):bs_type + ns(time_scaled,3):cc_t2db + ns(time_scaled,3):race + bs_type + sex + race + ethnicity + marital_status + patient_regional_location + age_at_bs + cc_t2db + cc_htn + cc_dyslip + cc_hf + cc_liver + (time_scaled | patient_id), REML=FALSE)
summary(model.lmer3)

# lmer for plot
model.lmer.plot <- lmer(data=df_bmi, bmi ~ glp1 + ns(time_scaled,3) + ns(time_scaled,3):glp1 + (time_scaled | patient_id), REML=FALSE)
summary(model.lmer.plot) # aic = 1403825.7 with at least 3

# lmer with time_1yr indicator (failed to converge)
model_lmer1.plot <- lmer(data=df_bmi, bmi ~ glp1 + ns(time_scaled, 3) + ns(time_scaled, 3):glp1 + (time_scaled + time_1yr |patient_id), REML=FALSE)
summary(model.lmer.plot) # aic =

# without glp1 indicator
model_lmer1_c <- lmer(data=df_bmi, bmi ~ glp1 + ns(time_scaled, 3) + ns(time_scaled, 3):glp1 + (time_scaled + time_1yr |patient_id) + bs_type + sex + race + age_at_bs + cc_t2db + cc_htn + cc_dyslip + cc_hf + cc_liver, REML=FALSE)

model_lmer1_c <- lmer(data=df_bmi, bmi ~ glp1 + ns(time_scaled, 3) + ns(time_scaled, 3):glp1 + (time_scaled + time_1yr |patient_id) + bs_type + sex + race + ethnicity + marital_status + patient_regional_location + age_at_bs + cc_t2db + cc_htn + cc_dyslip + cc_hf + cc_liver, REML=FALSE)

# without glp1
model_lmer1_c <- lmer(data=df_bmi, bmi ~ns(time_scaled, 3) + (time_scaled + time_1yr |patient_id) + bs_type + sex + race + ethnicity + marital_status + patient_regional_location + age_at_bs + cc_t2db + cc_htn + cc_dyslip + cc_hf + cc_liver, REML=FALSE)
summary(model_lmer1_c)

# stratified by the time-to-initation
table(df_bmi$time_to_glp1_cat2)

# stratifies by time_to_glp1_cat
model_lmer1_stratified <- lmer(data=df_bmi, bmi ~ time_to_glp1_cat + ns(time_scaled, 3) + ns(time_scaled, 3):time_to_glp1_cat + (time_scaled + time_1yr |patient_id), REML=FALSE)

# stratifies by covariates
model_lmer1_bs <- lmer(data=df_bmi, bmi ~ bs_type + ns(time_scaled, 3) + ns(time_scaled, 3):bs_type + (time_scaled|patient_id), REML=FALSE)

summary(model_lmer1_bs)

# for GLP-1 users
model_lmer1_glp1 <- lmer(data=df_bmi %>% filter(glp1 ==1), bmi ~ ns(time_scaled, 3) + (time_scaled |patient_id), REML=FALSE)
summary(model_lmer1_glp1)

# for non-users (model fail to converge)
model_lmer1_non <- lmer(data=df_bmi %>% filter(glp1 ==0), bmi ~ ns(time_scaled, 3) + (time_scaled |patient_id), REML=FALSE)
summary(model_lmer1_non)
```

According to the AIC value, I will go for linear mixed effect model with random intercept and slope, which has the smallest AIC values among several modeling options. (model_lmer1).
```{r}
## for total population 
# save predicted outcome (this prediction is without random effect)
df_bmi$pred1_fixed = predict(model.lmer.plot, re.form=NA)
# prediction with random effect
df_bmi$pred1_re = predict(model.lmer.plot, re.form=NULL)

# for stratified by the time-to-initation
#df_bmi$pred1_fixed2 = predict(model_lmer1_stratified, re.form=NA)
#df_bmi$pred1_re2 = predict(model_lmer1_stratified, re.form=NULL)

## for GLP-1 users
df_bmi_glp1$pred1_fixed = predict(model_lmer1_glp1, re.form=NA)
df_bmi_glp1$pred1_re = predict(model_lmer1_glp1, re.form=NULL)

## for non-users
df_bmi_non$pred1_fixed = predict(model_lmer1_non, re.form=NA)
df_bmi_non$pred1_re = predict(model_lmer1_non, re.form=NULL)
```
## point out the bottom value of the fixed line
```{r}
# for glp1 users
df_bmi_glp1 <- df_bmi %>% filter(glp1==1)
min_value <- min(df_bmi_glp1$pred1_fixed, na.rm = TRUE) # 21.15038
nadir_glp1 <- df_bmi_glp1 %>% filter(pred1_fixed < 35) %>% print()

#min_value <- min(df_bmi_glp1$pred1_fixed, na.rm = TRUE)
# <- df_bmi_glp1 %>% summarise(min_value = min(pred1_fixed, na.rm = Tmin_valueRUE))

df_bmi_non <- df_bmi %>% filter(glp1==0)
```

### Spaghetti Plots with linear model
```{r}
# random sampled 50 individuals
set.seed(123)
id_sel = sample(df_bmi$patient_id, 50)
id_sel_glp1 = sample(df_bmi$patient_id[df_bmi$glp1 ==1], 50)
id_sel_non = sample(df_bmi$patient_id[df_bmi$glp1 ==0], 50)

## total population
# make time_shift variable
df_bmi$time_shift = df_bmi$time + ifelse(df_bmi$glp1==0, -0.1, 0.1)

spagh_plot2 <- ggplot() +
  #geom_point(data=df_bmi[df_bmi$patient_id %in% id_sel, ], aes(x=time, y=bmi, group=glp1, color=glp1), alpha=0.2) +
  #geom_line(data=df_bmi[df_bmi$patient_id %in% id_sel, ], aes(x=time, y=bmi, group=patient_id, color=glp1), alpha=0.1) +
  geom_line(data=df_bmi, aes(x=time, y=pred1_fixed, group=glp1, color=glp1), linewidth=1, linetype = "solid") +
  theme_classic() +
  ggtitle("A. Post-surgical BMI with Linear Mixed Model Fit, GLP-1 users vs. non-users")  + 
  xlab("Time since Bariatric surgery (in year)") +
  ylab("BMI") + 
  scale_x_continuous(
    breaks = seq(0, max(df_bmi$time, na.rm = TRUE), by = 365), 
    labels = seq(0, max(df_bmi$time, na.rm = TRUE) / 365, by = 1)
    ) +
  ylim(10, 50) + 
  scale_color_manual(
  values = c("0" = "black", "1" = "blue"),
  labels = c("Non-uesrs", "GLP-1 users"),
  name = "GLP-1 initiation"
)
  
spagh_plot2
ggsave("trajectory_comparison_glp1_vs_non_spline.png", plot = spagh_plot2, width = 8, height = 5, dpi = 300, bg = "white")

## total population with stratification across the time-to-events
df_bmi$time_shift = df_bmi$time + ifelse(df_bmi$glp1==0, -0.1, 0.1)
# time_to_glp1_cat
spagh_plot2_stratified <- ggplot() +
  #geom_point(data=df_bmi[df_bmi$patient_id %in% id_sel, ], aes(x=time, y=bmi, group=glp1, color=glp1), alpha=0.2) +
  geom_line(data=df_bmi[df_bmi$patient_id %in% id_sel, ], aes(x=time, y=bmi, group=patient_id, color=glp1), alpha=0.1) +
  geom_line(data=df_bmi, aes(x=time, y=pred1_fixed2, group=time_to_glp1_cat, color=time_to_glp1_cat), linewidth=1, linetype = "solid") +
  theme_classic() +
  labs(x= "Time (in year)",
       y ="BMI",
       title = "B. Post-surgical BMI with Linear Mixed Model Fit, \nstratified by the timing of GLP-1 initiation") +
  scale_x_continuous(
    breaks = seq(0, max(df_bmi$time_n, na.rm = TRUE), by = 365), 
    labels = seq(0, max(df_bmi$time_n, na.rm = TRUE) / 365, by = 1)
    ) +
  scale_color_discrete(labels = c("Never used", "within 1 year", "in 1-2 year", "in 2-3 year", "in 3-4 year", "in 4-5 year", "in 5-6 year", "in 6-7 year", "after 7 year")) +
  guides(color = guide_legend(title="Timing of GLP-1 initiation"))

spagh_plot2_stratified

ggsave("trajectory_comparison_stratified.png", plot = spagh_plot2_stratified, width = 8, height = 5, dpi = 300, bg = "white")
```
## stratified by 3 categories, glp1 users vs. non-users
```{r}
# time_to_glp1_cat2
spagh_plot2_stratified <- ggplot() +
  geom_point(data=df_bmi[df_bmi$patient_id %in% id_sel, ], aes(x=time, y=bmi, group=glp1, color=glp1), alpha=0.2) +
  geom_line(data=df_bmi[df_bmi$patient_id %in% id_sel, ], aes(x=time, y=bmi, group=patient_id, color=glp1), alpha=0.1) +
  geom_line(data=df_bmi, aes(x=time, y=pred1_fixed2, group=time_to_glp1_cat2, color=time_to_glp1_cat2), linewidth=1, linetype=2) +
  theme_bw() +
  labs(x= "Time (in days)",
       y ="BMI",
       title = "Spaghetti plot for BMI with linear mixed model fit \nstratified by GLP-1's initiation timing") +
  scale_color_discrete(labels = c("Never used", "Early used (within 2y)", "Middle used (between 2-5y)", "Late used (after 5y)")) +
  guides(color = guide_legend(title="Time to GLP-1 initiation")) + 
  scale_color_manual(
  values = c("0" = "black", "1" = "blue"),
  labels = c("Not on GLP-1", "On GLP-1"),
  name = "GLP-1 Status"
)

spagh_plot2_stratified

## for GLP-1 users
spagh_plot2_glp1 <- ggplot() +
  geom_point(data=df_bmi[df_bmi$patient_id %in% id_sel_glp1, ], aes(x=time, y=bmi, group=glp1), color="blue", alpha=0.2) +
  geom_line(data=df_bmi[df_bmi$patient_id %in% id_sel_glp1, ], aes(x=time, y=bmi, group=patient_id, color=glp1), color ="blue", alpha=0.1) +
  geom_line(data=df_bmi_glp1, aes(x=time, y=pred1_fixed, group=glp1), color="blue", linewidth=1, linetype=2) +
  theme_bw() +
  ggtitle("Spaghetti plot for BMI with LMM Fit among GLP-1 users")  + 
  xlab("Time since surgery (in days)") +
  ylab("BMI")

spagh_plot2_glp1 

## for non-users
spagh_plot2_non <- ggplot() +
  geom_point(data=df_bmi[df_bmi$patient_id %in% id_sel_non, ], aes(x=time, y=bmi, group=glp1), color="black", alpha=0.2) +
  geom_line(data=df_bmi[df_bmi$patient_id %in% id_sel_non, ], aes(x=time, y=bmi, group=patient_id, color=glp1), color = "black",alpha=0.1) +
  geom_line(data=df_bmi_non, aes(x=time, y=pred1_fixed, group=glp1), color="black", linewidth=1, linetype=2) +
  theme_bw() +
  ggtitle("Spaghetti plot for BMI with LMM Fit among non-users")  + 
  xlab("Time (in days)") +
  ylab("BMI")

spagh_plot2_non
```

## incidence rate
```{r}
new_data <- data.frame(
  time_to_glp1_cat = as.factor(c(1,1,2,3,4,5,6,7,8)),
  time = c(0,365*1,365*2,365*3,365*4,365*5,365*6,365*7,365*8),
  glp1 = c(1,1,1,1,1,1,1,1,1),
  patient_id = NA  
)
# time = c(0,365*1,365*2,365*3,365*4,365*5,365*6,365*7,365*8), 
predicted_values <- predict(model_lmer1_stratified, newdata = new_data, re.form = NA)
result <- cbind(new_data, predicted_bmi = round(predicted_values, 5))
result 

result_table <- result %>%
  mutate(
    year = c(0, 1, 2, 3, 4, 5, 6, 7, 8),
    predicted_bmi = sprintf("%.2f", predicted_bmi)) %>%
  ggplot(aes(x = factor(year), y = 1, label = predicted_bmi)) +
  geom_text(size = 3.5, vjust = -0.3, hjust = 0.5) + # Adjust size, vertical, and horizontal position
  labs(title = "Predicted BMI at the time of GLP-1 initiation for each year") +
  theme_classic2() +
  theme(
    axis.title = element_blank(),
    axis.text.y = element_blank(),
    axis.text.x = element_blank(),
    panel.grid = element_blank(),
    plot.title = element_text(vjust = 0, hjust = 0.01, size =12)
  )
print(result_table)
```
## pooling the plot
```{r}
### combine
library(patchwork)
combined_plot <- (spagh_plot2_stratified/ result_table) + plot_layout(heights = c(3, 0.5, 0.1, 0.1))
print(combined_plot)

ggsave("trajectory_comparison_stratified_ir.png", plot = combined_plot, width = 7, height = 5, dpi = 300, bg = "white")
```

### Quantifying the difference

We want to quantify and make inference about the difference in population means of the two groups by plotting the difference of predicted negpan scores between the two groups as function of time, with corresponding 95% CI. 
```{r}
summary(model_lmer1_c)
summary(model_lmer1_c)$coefficients
summary(model_lmer1_c)$coefficients[6:8,1]

A=matrix(ns(c(1,365*1,365*2,365*3,365*4,365*5,365*6,365*7,365*8, 365*9),Boundary.knots=c(1,365*9),knots=c(365*1.5,365*4.5)),ncol=3)

#A=matrix(ns(c(1,365*1,365*2,365*3,365*4,365*5,365*6,365*7,365*8, 365*9),Boundary.knots=c(1,365*9),knots=c(365*1.5,365*4.5)),ncol=4)

b=matrix(summary(model_lmer1_c)$coefficients[6:8,1],ncol=1)
var_b=summary(model_lmer1_c)$vcov[6:8,6:8]

fit=A %*% b
var_fit = A %*% var_b %*% t(A)
se_fit = sqrt(diag(var_fit))

result=data.frame(time=c(1,365*1,365*2,365*3,365*4,365*5,365*6,365*7,365*8,365*9),fit=fit,se_fit=se_fit,ci_l=fit-2*se_fit,ci_u=fit+2*se_fit)

# plot results
ggplot(data=result,aes(x=time, y=fit)) + geom_point() + geom_line() +
  geom_abline(intercept=0,slope=0,col="red") + 
  geom_line(aes(x=time,y=ci_l),col="navyblue") +
  geom_line(aes(x=time,y=ci_u),col="navyblue") + 
  ylim(-3,5) + ylab("Estimated Mean Difference in Negative Symptom Score") + 
  xlab("Time (in weeks)") +
  ggtitle("Mean Negative Symptom Score Difference between Groups") +
  theme_bw()
```
##
```{r}
library(gtsummary)
library(dplyr)
head(df)
df %>% dplyr::select(age_at_bs, race, bs_type) %>% tbl_summary()

tbl_regression(model_lmer1)
```

# 2, Time_to_event analysis component
### preprocessing for time-to-event dataset
Model: Cox proportional hazard model
Data frame: one row per individual
```{r}
# generate 'study end date' 
df <- df %>%
  mutate(study_end_date = ymd("2023-12-31"))
# define exit_date
df <- df %>%
  mutate(
    glp1_initiation_date = as.Date(glp1_initiation_date),
    death_date = as.Date(death_date),
    study_end_date = as.Date(study_end_date), 
    glp1 = case_when(
      temporality == 0 ~ 0, 
      temporality == 2 ~ 1
    )
  )
df <- df %>%
  mutate(exit_date = pmin(glp1_initiation_date, death_date, study_end_date, na.rm = TRUE))
print(df$exit_date, 10)

# calculate time-to-exit [person-day]
df <- df %>%
  mutate(entry = 0, 
         time_to_exit = exit_date - bs_date)   # unit of time_to_exit = difftime [days]

# convert time_to_exit to numeric
df <- df %>%
  mutate(entry = 0, 
         time_to_exit = as.numeric(exit_date - bs_date), units = "days")

# any 0 or negative value of time-to-exit? yes, 35 
df_check <- df %>%
  filter(time_to_exit <=0) %>%
  summarise(negative_time = n()) %>%
  print() 

# find the problematic_rows
# 15 individuals have (death_date = bs_date) + 20 individuals have (glp1_initiation_date = bs_date)
problematic_rows <- df %>%
  filter(time_to_exit <= entry | is.na(time_to_exit) | is.na(entry)) %>%
  dplyr::select(patient_id, bs_date, glp1_initiation_date, death_date, study_end_date, time_to_exit, entry) %>%
  print()

# delete the 35 problematic_rows by using anti_join
df_clean <- df %>%
  anti_join(problematic_rows, by = "patient_id")

nrow(df_clean) # 37608

# number of study population # 17010 <- glp1 users = 6315 | non-users = 31293
unique_patients <- length(unique(df_clean$patient_id[df_clean$glp1 ==0]))
print(unique_patients)
```
### fit the COX model
```{r}
surv_obj <- Surv(time= df_clean$entry, time2= df_clean$time_to_exit, event= df_clean$glp1)
```

## Joint model
1. Clean the dataset for each model
2. Use a common subset of patients present in both datasets
```{r}
## 1. clean the dataset for each model
## 1-1. clean the data for 'df_cox' dataset
# can I just drop the col with missing - b/c I don't use them 
df_clean <- df_clean %>%
  dplyr::select(-bmi_index, -glp1_date, -glp1_initiation_date, -glp1_gap, -gap_glp1_bs, -death_year, -death_month, -death_date,-reason_yob_missing, -month_year_death, -death_date_source_id, -source_id, -year_of_birth, -encounter_id, -unique_id, -code_system, -code, -route, -brand, -strength, -derived_by_TriNetX, -Indication, -age_cat, -glp1_user, -temporality)
# check missingness
mice::md.pattern(df_clean, plot=FALSE)

#  Stop time must be > start time
names(df_clean)
check <- df_clean %>% filter(exit_date <= bs_date) %>% print()
nrow(is.na(check)) # 0

## 1-2. clean the data for 'df_lme' dataset
# change the format of time : timediff -> numeric
df_bmi$time <- as.numeric(df_bmi$time)
# to compare the time vs. time_to_exit, add time_to_exit to the df_lme
df_bmi_clean <- df_bmi %>%
  left_join(df_clean %>% dplyr::select(patient_id, time_to_exit), by = "patient_id")
nrow(df_bmi_clean) # 287164
df_bmi_clean <- df_bmi_clean  %>%
  filter(time < time_to_exit)
nrow(df_bmi_clean) # 283986

# 2, Use a common subset of patients present in both datasets:
common_ids <- intersect(unique(df_bmi_clean$patient_id), unique(df_clean$patient_id))
df_lme <- df_bmi_clean[df_bmi_clean$patient_id %in% common_ids, ] # long dataset 
df_cox <- df_clean[df_clean$patient_id %in% common_ids, ] # 1 row per 1 person

n <- length(unique(df_lme$patient_id)) # n = 15606
m <- length(unique(df_cox$patient_id)) # m = 15606
nrow(df_cox) # m = 15606

# Check for missing or infinite values
mice::md.pattern(df_cox, plot=FALSE) # 0
mice::md.pattern(df_lme, plot=FALSE) # 0

# the ordering of the subjects needs to be the same
head(df_cox)

exp(0.0071) #1.007125
exp(0.1*1.3944)

check <- df_lme %>%
  filter(patient_id == "6A_lD") %>% dplyr::select(patient_id, time_to_exit, time) %>% print()
```
#### Method 1 : "jointModel" function
solution : 
1. increase iteration
2. Adjust Convergence Criteria
```{r}
### LMM + COX model 
df_cox <- as.data.frame(df_cox)
df_lme <- as.data.frame(df_lme)

head(df_lme)
# longitudinal analysis (Increase Iteration Limits)
lmeFit0 <- lme(bmi ~ glp1 + ns(time, 3):glp1, random = ~ time | patient_id, data=df_lme, control = lmeControl(opt = "optim", msMaxIter = 100, niterEM = 50))

lmeFit1 <- lme(bmi ~ ns(time, 3), random = ~ time | patient_id, data=df_lme, control = lmeControl(opt = "optim", msMaxIter = 100, niterEM = 50))

lmeFit2 <- lme(bmi ~ ns(time, 3) + bs_type + sex + race +  ethnicity + marital_status + patient_regional_location + age_at_bs + cc_t2db + cc_htn + cc_dyslip + cc_hf + cc_liver, random = ~ time | patient_id, data=df_lme, control = lmeControl(opt = "optim", msMaxIter = 100, niterEM = 50))

lmeFit3 <- lme(bmi ~ ns(time, 3):bs_type + ns(time, 3):cc_t2db + ns(time, 3):race + bs_type + sex + race +  ethnicity + marital_status + patient_regional_location + age_at_bs + cc_t2db + cc_htn + cc_dyslip + cc_hf + cc_liver, random = ~ time | patient_id, data=df_lme, control = lmeControl(opt = "optim", msMaxIter = 100, niterEM = 50))


# time-to-event analysis
coxFit <- coxph(Surv(df_cox$time_to_exit, df_cox$glp1) ~ bs_type + sex + race +  ethnicity + marital_status + patient_regional_location + age_at_bs + cc_t2db + cc_htn + cc_dyslip + cc_hf + cc_liver, data=df_cox, x=TRUE, model=TRUE) 
summary(coxFit)

# joint model
## 1.under 
jointFit <- jointModel(lmeFit0, coxFit, timeVar = "time", method = "weibull-PH-aGH", control = list(niter = 200, eps = 1e-6)) # run
summary(jointFit)
jointFit1 <- jointModel(lmeFit1, coxFit, timeVar = "time") # run
summary(jointFit1)
```

```{r}
# with model 0
exp(0.0484) # 1.04959
0.0484 + 1.96*0.0021 # 0.052516
0.0484 - 1.96*0.0021 # 0.044284
exp(0.052516) # 1.053919
exp(0.044284) # 1.045279

# with model 1
exp(0.0018) # 1.001802 -> 0.2%
0.0018 + 1.96*0.0027 # 0.007092
0.0018 - 1.96*0.0027 # -0.003492
exp(0.007092) # 1.007117
exp(-0.003492) # 0.9965141

exp(0.0071) # aHR = 1.007125 (p<0.001)
exp(0.0027) # SE = 1
0.0071 + 1.96*0.0027 # 0.012
0.0071 - 1.96*0.0027 # 0.0018

exp(0.012) # 1.012
exp(0.0018) # 1.0018

exp(0.0029) # 1.002904
0.0029 + 1.96 * 0.0029 # 0.008584
0.0029 - 1.96 * 0.0029 # -0.002784

exp(0.008584) # 1.008621
exp(-0.002784) # 0.997219

  
  
exp(0.0748) # 1.077669
exp(0.0024) # 1.002403

0.0748 + 1.96 * 0.0024	# 0.079504
0.0748 - 1.96 * 0.0024	# 0.070096
exp(0.079504) # 1.08275
exp(0.070096) # 1.072611



## 2~3. under 'Cox-PH-GH' assumption
jointFit2 <- jointModel(lmeFit, coxFit, timeVar = "time")
# error: Error in model.frame.default(data = FALSE, formula = ~long + W1 + W2,  : 'data' must be a data.frame, environment, or list
summary(jointFit2)
exp(0.0011) # aHR = 1.001101 

jointFit3 <- jointModel(lmeFit, coxFit, timeVar = "time", method = "Cox-PH-GH", control = list(iter.EM = 500, iter.qN = 500, tol1 = 1e-6, tol2 = 1e-6, verbose = FALSE))

# 4. use another package (more recent one)
if (!require("JMbayes2")) install.packages("JMbayes2")
library(JMbayes2)
jointFit4 <- jm(coxFit, lmeFit, time_var = "time") # error with long statement

### 5~6. survival analysis with Weibull 
install.packages("flexsurv")
library(flexsurv)
weibullFit <- survreg(Surv(df_cox$time_to_exit, glp1) ~ sex + cc_obs, data=df_cox, x=TRUE)
summary(weibullFit)

jointFit5 <- jointModel(lmeFit, weibullFit, timeVar = "time")
jointFit6 <- jointModel(lmeFit, weibullFit, timeVar = "time", method = "weibull-PH-aGH", control = list(niter = 200, eps = 1e-6))

mice::md.pattern(weibullFit, plot=FALSE)
```

### plot JointModel with Conditional survival predictions
```{r}
# plot the conditional survival plot
?survfitJM

df_lme_0gHN <- df_lme[df_lme$patient_id == "0gHN", ]
lapply(df_lme_0gHN, function(x) if (is.factor(x)) levels(x))

# 1
sfit <- survfitJM(jointFit, newdata = df_lme_0gHN[1:5, ], idVar = "patient_id")

# 2
sfit <- survfitJM(jointFit, newdata = df_lme_0gHN, idVar = "patient_id",  Thoriz = NULL, Dt = NULL, simulate = FALSE)

# 3
sfit <- survfitJM(jointFit, newdata = df_lme_0gHN, idVar = "patient_id", survTimes = 365*2, last.time = 365*5, simulate = FALSE)

sfit
plot(sfit, include.y = TRUE)

## Time-Dependent AUCs for Joint Models
#aucJM(jointFit, newdata = df_lme, idVar = "patient_id", Tstart = 365, Thoriz = 365*7)
```
### plot JointModel with KM curve
```{r}
# plot 
plot(jointFit, 3, add.KM = TRUE, col = "red", lwd = 2)

par(mfrow = c(2, 2))
plot(jointFit)
```
#### Method 3 : Adding time-dependent slopes
```{r}
# longitudinal analysis (Increase Iteration Limits)
lmeFit <- lme(bmi ~ ns(time, 3), random = ~ time | patient_id, data=df_lme, control = lmeControl(opt = "optim", msMaxIter = 100, niterEM = 50))

# time-to-event analysis
coxFit <- coxph(Surv(df_cox$time_to_exit, df_cox$glp1) ~ sex + cc_obs, data=df_cox, x=TRUE, model=TRUE) 

#joint model with slope
jointFit_slope <- jm(coxFit, lmeFit, time_var = "time",
                    functional_forms = ~ slope(bmi, eps =1, direction = "back"))
?jm

summary(jointFit_slope)
```
### 2 stage model
```{r}
# Stage 1: Fit the Longitudinal Submodel
### longitudinal analysis (Increase Iteration Limits)
lmeFit <- lme(bmi ~ ns(time_n, 3), random = ~ time_n | patient_id, data=df_lme, control = lmeControl(opt = "optim", msMaxIter = 100, niterEM = 50))

# extract subject-specific predicted value from the longitudinal model
df_lme_2stage <- df_lme
df_lme_2stage$pred1_fixed = predict(lmeFit, re.form=NA)

# method 1 : mean value
aggregated_data_1 <- df_lme_2stage %>%
  group_by(patient_id) %>%
  summarize(predicted_bmi = mean(pred1_fixed))

df_cox_1 <- df_cox %>%
  left_join(aggregated_data_1, by = "patient_id")

# Stage 2: Fit the survival model using predicted values from the longitudinal model
cox_2stage_1 <- coxph(Surv(time_to_exit, glp1) ~ predicted_bmi + sex + cc_obs, data = df_cox_1) 
summary(cox_2stage_1) # aHR = 1.07448, (1.067, 1.082)


## method 2: Last Observed Value
aggregated_data_2 <- df_lme_2stage %>%
  group_by(patient_id) %>%
  summarize(predicted_bmi = pred1_fixed[which.max(time_n <= time_to_exit)])

df_cox_2 <- df_cox %>%
  left_join(aggregated_data_2, by = "patient_id")

# Stage 2: Fit the survival model using predicted values from the longitudinal model
cox_2stage_2 <- coxph(Surv(time_to_exit, glp1) ~ predicted_bmi + sex + cc_obs, data = df_cox_2) 
summary(cox_2stage_2) # aHR = 1.0632, (1.056, 1.07)

```

```{r}
### Extract subject-specific random effects
random_effects <- ranef(lmeFit)

random_effects <- as.data.frame(random_effects)
random_effects$patient_id <- rownames(random_effects)
random_effects
nrow(random_effects) # 15606
nrow(df_cox) # 15606

random_effects <- random_effects %>%
  dplyr::select(-group, -term, -patient_id)

random_effects <- random_effects %>%
  rename(patient_id = level)

df_cox_2 <- df_cox %>%
  left_join(random_effects, by = "patient_id")
```





